OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-03 main.tex
Finished the text.

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype.  A and
B. Average changes compared to raw on the T47D and K562 sets.  The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices.  All
results in this figure are based on Pearson correlations between the
observed over expected counts.  Our method increases reproducibility of
similar samples while being able to discriminate different ones.}
\label{fig:aberrant}
\end{figure}

@@

Regarding cells with diploid genomes, we confirmed that our bias
correction strategy also achieves good results in terms separation between
unrelated and similar samples (Figure \ref{fig:normal} and Supplementary
Figures 6 and 7), increasing the reproducibility across different
experimental protocols for primary B and ES mouse cell samples. Again,
Spearman correlation of contacts revealed as the worst metric.

\begin{figure}
	\centerline{\includegraphics[width=.50\textwidth]{img/correlation_normal_figure4.pdf}}
	\caption{
		Results of the comparison between samples with normal karyotype. A Average difference and 95\% CI between each correction method (Y axis) and the raw data (brown dashed line) for the mm10 sets. B Corresponding ROC curves and AUCs. The introduced strategy does not under perform compared to the alternatives.
	}\label{fig:normal}
\end{figure}

\subsection{Computing resources}

One of the main reasons for the broad usage of the iterative correction \citep{imakaev2012iterative} is the easy and fast implementation, moreover compared to existing explicit methods \citep{servant2012hitc}. Our proposal uses only the total number of contacts per bin instead the full matrix for the model estimation. Thus, it achieves similar computing times that the fastest approaches (Figure \ref{fig:times} A). Yet it scales with increasing resolutions (decreasing bin size) (Figure \ref{fig:times} B).

\begin{figure}
	\centerline{\includegraphics[width=.35\textwidth]{img/times_sina_figure5.pdf}}
	\caption{
		Computing time (Y axis, log scale) of all the bias correction methods used. Each dot corresponds to one sample. The only method faster than ours under performs in all sample comparisons.
	}\label{fig:times}
\end{figure}

@@

@@

We would like to thank all members of the four labs involved in the 4DGenome Synergy project for the fruitful discussions during project meetings. EV wants to acknowledge specially the members of Miguel Beato's lab for their insights during lab meetings.

@@

This work was supported by the Spanish Ministry of Economy and
Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
SEV-2012-0208, ERC Synergy Grant 609989 (GF) and...
\vspace*{-12pt}

>>>>

%% --------------- Introduction --------------- %%

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on Pearson correlations between the
observed over expected counts.}
\label{fig:aberrant}
\end{figure}

@@

Does the performance of OneD with aberrant karyotypes comes at the cost of
decreased performance with normal karyotypes? To address this question, we
assembled another data set comprised of normal mouse B cells and embryonic
stem (ES) cells. The cell types were pooled in almost equal proportions
(see Table~\ref{tab:samples}) and the same tests as above were performed.

In these conditions, variations of the protocol had a strong effect on the
impact of the differen normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caib} and \textit{ice} increased the similarity when
the protocols were different, but decreased it when the protocols were the
same. The effect was stronger when comparing identical cell types, but the
same trend appeared when compaing different cell types, indicating that
these methods may enhance or reduce biological variation, depending on the
context. Once more, \textit{vanilla} followed the same trend as
\textit{ice} to a lesser extent. The \textit{lgf} method increased the
similarity when comparing the same cells with different experimental
protocols, and had little to no effect in the other cases. This indicates
that \textit{lgf} is a safe choice in this case.

OneD decreased the similarity between different cell types when using the
same protocol and increased it between identical cell types when using
different protocols. In this other two cases, it had little effect. In
summary, OneD never enhanced the experimental noise and it reduced it in
one more case than \textit{lgf}.

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
pairs of biologicate pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
on this problem, even though it was not always above the others methods at
every position of the graph.

Using other metrics to compare matrices gave similar results
(Supplementary Figures 6 and 7). OneD was always among the top scoring
methods. In these conditions, Spearman correlation of contacts again
appeared as as the worst comparison metric as it showed a lower
performance for all the normalization methods.

Taken togetheer, these results indicate that OneD performs as well as the
best normalization methods even with normal karyotypes.

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]{img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotype.  A.
Average changes compared to raw on the mouse data set. The bars are as in
B. ROC curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

\subsection{Speed}

Finally, we compared the computational speed of the different
normalization methods. One of the main reasons for the broad acceptance of
\textit{vanilla} and \textit{ICE} as standards is the speed of the
available implementation \citep{imakaev2012iterative}. This is even 
more significant that current explicit methods \citep{servant2012hitc} are
much slower in comparison.

Unlike those, OneD corrects a single variable instead of the whole matrix.
It uses only the total number of contacts per bin and thus estimates the
model parameters much faster than previous explicit methods.
Figure\ref{fig:times} shows the running times of the different methods on
the samples described in Table~\ref{tab:samples}. The fastest method is
\textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
span between the two.

OneD is the second fastest method and it always ran in less than 1 min in
the conditions of the benchmark. Interestingly, the running time of OneD
is much more homogeneous than that of the other methods. The reason is
that the size of the regression problem to be solved by the \texttt{mgcv}
package is always the same at fixed reslution.

Taken together, these results show that OneD is very competitive in terms
of computational speed compared to existing methods.

\begin{figure}
\centerline{\includegraphics[width=.35\textwidth]{img/times_sina_figure5.pdf}}
\caption{
Computing time of the bias correction methods. Each dot corresponds to
one sample. The only method faster than ours under performs in all sample
comparisons. Note the logarithmic scale on the y-axis.}
\label{fig:times}
\end{figure}

%% --------------- Discussion --------------- %%

@@

Here we introduced OneD, a fast computational method to normalize Hi-C
matrices. OneD was developed ground up to address the need to normalize
data collected on samples with aberrant karyotypes, but it applies
seemlessly to the case of normal karyotypes. We showed that OneD performs
significantly better than other methods when the cells present karyotypic
aberrations (Figure~\ref{fig:aberrant}), and that it performs equall well
as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
that OneD is on average faster than \textit{ice}, which makes it very
competitive from the point of view of computational speed.

The originality of OneD lies in that it projects all the biases onto a
single variable: the total amount of contact per bin. This allows greater
running speed, while preserving a good performance on samples with
karyotypic aberrations. One of the reasons that OneD is better able to
highlight the biological distinctions between samples is that it does not
correct the copy number but enhances it (Figures~\ref{fig:totals} and
\ref{fig:copy_number}).

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. Note that in the presence of
translocations, deletions and large scale rearrangements, correcting the
copy number is equally faulty as not correcting it. Indeed, in this case
the reference genome on which the matrix is drawn is incorrect and
changing the values of the pixels cannot account for this information.

Depending on the intention of the user, the effect of the copy number
should either be kept or removed. This is why OneD leaves does not perform
the correction by default, but allows the user to obtain a
diploid-equivalent Hi-C map computed from a hidden Marko model. The
resulting matrices have a near constant amount of contacts per bin, but
the artifacts caused by the misfit between the genome of the sample and
the reference genome are still present (for instance, the artifacts caused
by large scale inversions are not changed in any way).

Overall, OneD constitutes a safe choice to normalize Hi-C matrices. If the
karyotype of the sample is aberrant, it enhances the biological variation.
If not, it performs at least equally well as other methods with an
advantage in terms of computational speed.

@@

We would like to thank the members of the 4DGenome Synergy project for the
fruitful discussions during project meetings. EV wants to acknowledge
the members of Miguel Beato's laboratory for their insights during lab
meetings.

@@

This work was supported by the Spanish Ministry of Economy and
Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
SEV-2012-0208, ERC Synergy Grant 609989 and...

\vspace*{-12pt}
